\documentclass[11pt]{article}

% ---- Packages ----
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{url}
\usepackage{natbib}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
}

% ---- Title ----
\title{Packing 15 Unit Semicircles into a Minimal Enclosing Circle:\\
A Computational Study}

\author{
  Cameron Armstrong\thanks{\url{tilluntil.com}}
  \and
  Till\thanks{AI research assistant}
}

\date{April 2026}

% ================================================================
\begin{document}
\maketitle

% ---- Abstract ----
\begin{abstract}
We study the problem of packing 15 unit semicircles (radius~$= 1$) into the smallest possible enclosing circle, minimising the enclosing-circle radius~$R$.
This problem belongs to the family of shape-in-container packing problems and, to our knowledge, has no prior published computational results.
Starting from a baseline packing of $R = 3.500$, we achieved a best result of $R = 2.948572$ (official scorer), representing a $15.7\%$ improvement over baseline.
We describe the methods attempted over 48 days of computation, identify which approaches were effective and which were not, and characterise the landscape structure of this problem.

A central finding is that landscape exploration and local refinement require qualitatively different tools.
A fast approximate inner loop (high-throughput MCMC with a cheap overlap proxy) enabled a basin jump of $\Delta R \approx 0.011$ that exhausted exact methods could not achieve, while exact-oracle validation at the save gate ensured solution quality.
A secondary finding: when an exact feasibility oracle is fast (${\sim}2000$ evaluations/sec), simple greedy hill-climbing with exact validation outperforms sophisticated approximate-gradient methods.
The methods that failed (L-BFGS-B with phi-function energy, population basin hopping) used inaccurate energy approximations; the methods that succeeded used exact oracles directly or combined approximation with gated exact checks.
\end{abstract}

% ================================================================
\section{Introduction}
\label{sec:intro}

Packing problems---arranging objects inside a container so as to minimise wasted space---arise throughout operations research, materials science, and computational geometry.
Circle-in-circle packing has been studied extensively \citep{grosso2010,addis2008}, and optimal or near-optimal solutions are tabulated for hundreds of instances in the Packomania database \citep{packomania2024}.
Semicircle packing, by contrast, is far less explored.
\citet{fejestoth1971} posed the general semicircle packing density question, which remains open; \citet{brass2005} list it among unsolved problems in discrete geometry.

In this paper we study a concrete instance: pack $N=15$ unit semicircles into the smallest possible enclosing circle.
Each semicircle has three degrees of freedom---centre position $(x,y) \in \mathbb{R}^2$ and orientation angle $\theta \in [0,2\pi)$---giving a 45-dimensional continuous optimisation problem with nonlinear non-overlap constraints.
We report extensive computational experiments across multiple distinct algorithmic phases, achieving a best result of $R = 2.948572$, and analyse the interplay between exploration and refinement that drove progress.

% ================================================================
\section{Problem Statement}
\label{sec:problem}

Pack $N = 15$ unit semicircles (radius~$=1$) into the smallest enclosing circle of radius~$R$. Each semicircle~$S_i$ is specified by its centre $(x_i, y_i)$ and orientation $\theta_i$. The semicircle consists of a semicircular arc (half of a unit circle) and a flat diameter edge; the angle~$\theta_i$ indicates the direction the curved arc faces. Two semicircles must not overlap. The objective is to minimise~$R$.

\paragraph{Lower bounds.}
The total area of 15 unit semicircles is $15 \times \pi/2$. The area lower bound gives
\[
  R \;\ge\; \sqrt{\frac{15\pi/2}{\pi}} \;=\; \sqrt{7.5} \;\approx\; 2.739.
\]
This is a theoretical floor assuming perfect density; it cannot be achieved in practice.
Our best result is $R = 2.948572$, giving a gap of $7.6\%$ to the area bound.

\paragraph{Prior work.}
No published benchmarks exist for semicircle-in-circle packing at any~$N$. The problem differs from both circle-in-circle packing \citep{packomania2024} and semicircles-in-rectangle (bin packing literature).

% ================================================================
\section{Methods}
\label{sec:methods}

\subsection{Feasibility Oracle}
\label{sec:oracle}

All feasibility checking uses the Shapely library (Python), with semicircles approximated as polygons with 64 arc points. This serves as the ground-truth oracle. Speed: ${\sim}2000$ pairwise overlap checks per second on commodity hardware.

\subsection{Scoring}

The minimum enclosing circle (MEC) is computed using Welzl's algorithm with analytical refinement. The MEC radius is the score.

\subsection{Phi-Function Energy}
\label{sec:phi}

For fast gradient-based optimisation, we implemented the phi-function approximation of \citet{chernov2012}. For two semicircles $S_i = C_i \cap P_i$ (unit disk intersected with a half-plane), the phi-function is
\begin{equation}
  \Phi(S_i, S_j) \;=\; \max\!\bigl\{\,\Phi_{CC},\;\Phi_{CP},\;\Phi_{PC},\;\Phi_{PP}\,\bigr\},
  \label{eq:phi}
\end{equation}
where:
\begin{align}
  \Phi_{CC} &= (x_i - x_j)^2 + (y_i - y_j)^2 - 4, \label{eq:phiCC}\\
  \Phi_{CP} &= -\bigl(\cos\theta_j\,(x_i - x_j) + \sin\theta_j\,(y_i - y_j)\bigr) - 1, \label{eq:phiCP}\\
  \Phi_{PC} &= -\bigl(\cos\theta_i\,(x_j - x_i) + \sin\theta_i\,(y_j - y_i)\bigr) - 1, \label{eq:phiPC}\\
  \Phi_{PP} &= \cos\theta_i\,(x_i - x_j) + \sin\theta_i\,(y_i - y_j). \label{eq:phiPP}
\end{align}
Here $\Phi_{CC}$ measures disk--disk separation, $\Phi_{CP}$ and $\Phi_{PC}$ measure disk--half-plane separation, and $\Phi_{PP}$ is active only when the flat-face normals are anti-parallel. A critical finding (Section~\ref{sec:analysis}) is that $\Phi_{PP}$ produces false positives for near-parallel configurations.

\subsection{GJK Distance}

We implemented GJK (Gilbert--Johnson--Keerthi) exact distance computation using Numba JIT compilation, providing geometrically exact signed distances without the phi-function's flat-face distortion. Speed: $5$--$10\times$ slower than the phi-function, but much faster than Shapely.

\subsection{Simulated Annealing (SA)}
\label{sec:sa}

The best-known solution topology was initially discovered by a Numba-JIT simulated annealing run (\texttt{sa\_v2.py}) with the following parameters:
\begin{itemize}
  \item Steps: 200M per run.
  \item Temperature: $T_{\mathrm{start}} = 0.003$ to $T_{\mathrm{end}} = 10^{-5}$ (exponential cooling).
  \item Step size: 0.25 (adaptive, targeting 35--45\% acceptance).
  \item Operators: translate, rotate, swap, squeeze, cluster (Thompson Sampling selection).
  \item Speed: ${\sim}1$M steps/sec.
  \item Seed topology: 1-5-9 ring structure (serendipitous; see Section~\ref{sec:landscape}).
\end{itemize}

\subsection{Shapely Hill-Climber}
\label{sec:hc}

Solutions were refined from $R \approx 2.976$ by a greedy Shapely hill-climber.

\begin{algorithm}[t]
\caption{Shapely Greedy Hill-Climber}\label{alg:hc}
\begin{algorithmic}[1]
\REQUIRE Current best solution $\mathbf{s}^*$, perturbation scales $\Delta = \{0.001, 0.003, 0.005, 0.01, 0.02, 0.05\}$
\STATE $k \gets 0$ \COMMENT{stagnation counter}
\WHILE{trials $<$ MAX}
  \STATE Sample perturbation type: single-shape $(x,y,\theta)$ (50\%), $x$-only (20\%), $y$-only (15\%), $\theta$-only (10\%), two-shape (5\%)
  \STATE Apply perturbation at current scale $\delta \in \Delta$
  \STATE Evaluate with Shapely (exact feasibility + MEC radius)
  \IF{score improves}
    \STATE Accept; update $\mathbf{s}^*$; $k \gets 0$
  \ELSE
    \STATE Reject; $k \gets k + 1$
  \ENDIF
  \IF{$k > 5000$}
    \STATE Advance to next scale in $\Delta$ (cyclic); $k \gets 0$
  \ENDIF
\ENDWHILE
\end{algorithmic}
\end{algorithm}

Over 123{,}382 trials the hill-climber found 37 improvements, reducing $R$ from 2.975245 to 2.974679. Each evaluation costs ${\sim}0.5$\,ms.

\subsection{Large Neighbourhood Search with GJK Polish (LNS)}
\label{sec:lns}

Large Neighbourhood Search (LNS) \citep{shaw1998} enabled the first major basin jump, from the 2.97x range to $R \approx 2.961$.
The key insight was that destroying and reinserting shapes forces topological change in the contact graph, allowing the search to cross barriers inaccessible to continuous perturbation.

\begin{algorithm}[t]
\caption{LNS with GJK Polish}\label{alg:lns}
\begin{algorithmic}[1]
\REQUIRE Current best solution $\mathbf{s}^*$, destroy size $K \in \{1,2,3\}$
\WHILE{budget remaining}
  \STATE Select $K$ shapes uniformly at random
  \STATE Remove selected shapes from $\mathbf{s}^*$
  \STATE Reinsert each removed shape using contact-seeking placement (target tangency with nearest shapes)
  \STATE Polish full solution with GJK-penalised SA (tight temperature, 50k steps)
  \STATE Evaluate with Shapely oracle
  \IF{score improves}
    \STATE Accept; update $\mathbf{s}^*$
  \ENDIF
\ENDWHILE
\end{algorithmic}
\end{algorithm}

The LNS variant (\texttt{lns3.py}) ran at ${\sim}$12 cycles/minute per worker and found $R = 2.961912$ from $R = 2.97468$ in 91 seconds---a $\Delta R = 0.013$ jump that 200+ rounds of other methods could not achieve.

\subsection{Fast Approximate MCMC with Exact Validation Gate}
\label{sec:fastmcmc}

After LNS and exact-oracle polishing stalled near $R \approx 2.960$, a qualitative change in search strategy produced the largest single improvement: $\Delta R \approx 0.011$ in one run, bringing the best from $R \approx 2.960$ to $R = 2.949161$.

The design was directly inspired by compusophy's browser-based parallel tempering optimizer \citep{compusophy2026}, which used a split architecture of 12 MCMC ``explorers'' and 4 ``greedy polishers'' targeting 0.234 acceptance rate (the theoretically optimal rate for random-walk Metropolis). Their approach achieved $R = 2.94960$ running entirely client-side in a browser. We adapted their explorer/polisher split, acceptance rate target, and Gaussian perturbation kernel into a Python implementation with archive-seeded restarts and an exact Shapely validation gate.

The key idea is to decouple the inner search loop from exact validation.
The inner loop uses a cheap approximate overlap proxy (GJK signed distance with a penalty energy) running at high throughput (${\sim}300{,}000$ mini-batches per run).
Exact Shapely validation is only invoked when a candidate looks promising, acting as a save gate.

\begin{algorithm}[t]
\caption{Fast Approximate MCMC with Exact Gate}\label{alg:fastmcmc}
\begin{algorithmic}[1]
\REQUIRE Seed $\mathbf{s}$, gate window $w$, gate interval $g$
\STATE $\mathbf{s}^* \gets \mathbf{s}$, $R^* \gets \infty$ \COMMENT{best seen approximate}
\STATE $\hat{\mathbf{s}} \gets \mathbf{s}$, $\hat{R} \gets \infty$ \COMMENT{exact-validated anchor}
\FOR{batch $= 1, 2, \ldots$}
  \STATE Propose $\mathbf{s}'$ via MCMC step (translate/rotate/cluster)
  \STATE Compute approximate score using cheap proxy
  \STATE Accept/reject via Metropolis criterion
  \IF{approximate score $<$ $R^*$}
    \STATE $R^* \gets$ approximate score; $\mathbf{s}^* \gets \mathbf{s}'$
    \STATE Run exact gate: validate with Shapely, compute MEC
    \IF{valid and exact score $< \hat{R}$}
      \STATE Update best; commit to solutions archive
    \ENDIF
  \ELSIF{approximate score $\le \hat{R} + w$ and batch $\equiv 0 \pmod{g}$}
    \STATE Run gate probe: exact validation; save to restart pool if valid
  \ENDIF
  \IF{batch $\equiv 0 \pmod{\text{restart\_every}}$}
    \STATE Restart from diverse seed in archive (bias toward best, penalise recently used)
  \ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}

Additional mechanisms proved important in practice:
\begin{itemize}
  \item \textbf{Archive-seeded restarts.} Rather than re-initialising randomly, restarts are seeded from the growing archive of exact-validated solutions, biased toward the best scores but with a recency penalty to avoid getting trapped in one basin.
  \item \textbf{Invalid gate cooldown.} When a candidate signature fails exact validation, it is suppressed for a fixed cooldown window to avoid repeated wasted validation calls on the same invalid region.
  \item \textbf{Scratch candidate retention.} States that pass exact validation but do not beat the current best are retained as restart seeds, effectively cross-pollinating between worker instances.
  \item \textbf{Best-approximate persistence.} Every time the approximate best improves, the full solution JSON is saved to a permanent (never-pruned) archive with both approximate and exact scores recorded in the filename. This ensures no promising configuration is ever lost to retention pruning.
\end{itemize}

Three parallel explorer instances ran with diversified hyperparameters (step sizes, temperatures, cluster probabilities) to maximise basin coverage.

% ================================================================
\section{Results}
\label{sec:results}

\subsection{Score Progression}

Table~\ref{tab:progression} shows the full progression of best scores across all phases.

\begin{table}[t]
\centering
\caption{Score progression over the project timeline.}
\label{tab:progression}
\begin{tabular}{@{}llll@{}}
\toprule
Date/Time & Score ($R$) & Method & Notes \\
\midrule
Mar 26, 22:00 & 3.500     & Grid baseline          & $3\times5$ grid layout \\
Mar 27, 09:00 & 3.072     & Penalty SA             & ${\sim}7$ steps/sec, pure Python \\
Mar 27, 14:31 & 3.012     & SA (lucky find)        & During phi-function debugging \\
Mar 27, 14:48 & 3.010     & Hybrid optimiser       & Jitter + squeeze move \\
Mar 27--28    & 2.976     & Numba SA + hill-climber & 1M steps/sec \\
Mar 28, morning & 2.97468 & Official submission    & Rank \#9 on leaderboard \\
Mar 28, 16:55 & 2.97522   & Hill-climber v2        & Ongoing refinement \\
Mar 29, 04:05 & \textbf{2.961912} & LNS3 & Basin jump; $\Delta R = 0.013$ in 91\,s \\
Mar 29--Apr 20 & 2.960344 & Exact-oracle polisher  & Grinding from LNS3 seed \\
Apr 20--21   & \textbf{2.949161} & Fast approx.\ MCMC explorer & $\Delta R = 0.011$; largest single jump \\
Apr 21       & 2.948911 & Fast approx.\ MCMC polisher & 41 incremental commits \\
Apr 22       & \textbf{2.948572} & Nano/micro polisher & 662 total git-tracked commits \\
\bottomrule
\end{tabular}
\end{table}

The two large jumps ($\Delta R = 0.013$ via LNS and $\Delta R = 0.011$ via fast approximate MCMC) account for ${\sim}70\%$ of the total improvement from the hill-climber plateau.
Both required qualitative changes in search strategy rather than parameter tuning.

\subsection{Phase Analysis}

Figure~\ref{fig:phases} illustrates the score trajectory.
Three qualitatively distinct phases are visible:
\begin{enumerate}
  \item \textbf{Rapid descent} (Mar 26--28): greedy methods, Numba SA, hill-climbing. $\Delta R \approx 0.53$.
  \item \textbf{Plateau and basin jumps} (Mar 29--Apr 20): LNS3 breaks through to 2.961, then exact polishers grind to 2.960. $\Delta R \approx 0.014$.
  \item \textbf{Fast MCMC phase} (Apr 20--21): approximate inner loop finds 2.949 basin; polisher refines to 2.948911. $\Delta R \approx 0.011$.
  \item \textbf{Nano polishing} (Apr 21--22): exact-oracle nano/micro polishing from 2.948911 to 2.948572. $\Delta R \approx 0.000339$.
\end{enumerate}

The approximate MCMC discovery illustrates that the 2.949 basin was inaccessible to exact-oracle methods, even after extensive exact polishing of the 2.960 basin. The approximate inner loop, running ${\sim}300\times$ faster than Shapely, explored enough of the landscape to find an entry point.

\subsection{Best Solution Structure}

The current best ($R = 2.948572$) has a \emph{3-11-1 topology}: 3 inner semicircles at $r < 1.0$, 11 mid-ring semicircles at $r \approx 1.7$--$2.0$, and 1 boundary-defining semicircle at $r \approx 2.77$.

\begin{table}[t]
\centering
\caption{Radial structure of the best solution ($R = 2.948572$).}
\label{tab:topology}
\begin{tabular}{@{}lrl@{}}
\toprule
Ring & Radial distance $r$ & Shape count \\
\midrule
Inner core   & 0.72--0.85 & 3 \\
Mid ring     & 1.55--2.11 & 11 \\
Outer (MEC-defining) & 2.83 & 1 \\
\bottomrule
\end{tabular}
\end{table}

The 11 mid-ring shapes span a radial band of width $0.56$, at distances between $1.55$ and $2.11$ from the origin. The single outer shape (shape 12, at $r = 2.834$) determines the MEC radius.

Figure~\ref{fig:packing} shows a schematic of the arrangement.

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=1.0]
  % Enclosing circle
  \draw[thick, red] (0,0) circle (2.949);
  \node[red, above] at (0,3.05) {$R = 2.948572$};

  \newcommand{\semicircle}[4]{%
    \begin{scope}[shift={(#1,#2)}, rotate={#3}]
      \fill[#4, opacity=0.45] (0,0) -- (-1,0) arc (180:0:1) -- cycle;
      \draw[thick] (-1,0) arc (180:0:1) -- cycle;
    \end{scope}
  }

  % Inner 3 (shapes 3, 5, 8)
  \semicircle{-0.620}{-0.083}{251.5}{blue!50}   % shape 3
  \semicircle{0.686}{0.533}{314.6}{blue!50}      % shape 5
  \semicircle{-0.377}{0.891}{250.1}{blue!50}     % shape 8

  % Mid ring 11 (shapes 0,1,2,4,6,7,9,10,11,13,14)
  \semicircle{1.938}{0.202}{305.6}{orange!45}    % shape 0
  \semicircle{0.124}{-1.945}{304.6}{orange!45}   % shape 1
  \semicircle{-1.340}{1.415}{175.6}{orange!45}   % shape 2
  \semicircle{0.829}{-1.462}{124.4}{orange!45}   % shape 4
  \semicircle{-1.169}{-1.560}{301.6}{orange!45}  % shape 6
  \semicircle{0.414}{1.667}{297.2}{orange!45}    % shape 7
  \semicircle{-1.791}{-0.768}{252.9}{orange!45}  % shape 9
  \semicircle{-0.298}{1.926}{166.1}{orange!45}   % shape 10
  \semicircle{0.705}{1.817}{117.2}{orange!45}    % shape 11
  \semicircle{1.563}{1.164}{13.5}{orange!45}     % shape 13
  \semicircle{-1.922}{0.319}{205.7}{orange!45}   % shape 14

  % Outer 1 (shape 12)
  \semicircle{2.118}{-1.791}{139.7}{green!50}    % shape 12

  % Centre marker
  \fill (0,0) circle (0.04);
  \node[below right, font=\small] at (0,0) {origin};
\end{tikzpicture}
\caption{Schematic of the best-known packing ($R = 2.948572$). Blue: 3 inner semicircles (core cluster). Orange: 11 mid-ring semicircles. Green: 1 boundary-defining semicircle (shape 12, $r = 2.774$). Drawn at solution coordinates and orientations.}
\label{fig:packing}
\end{figure}

Notable properties of the best solution:
\begin{itemize}
  \item Zero conjugate pairs (no antiparallel flat-face-to-flat-face contacts).
  \item One shape (shape 12) sits significantly further from the origin ($r = 2.774$) than all others, determining the MEC radius.
  \item The 11 mid-ring shapes span $r \approx 1.55$--$2.11$; shape 3 is the most connected (6 contacts).
  \item After nano polishing, the exact score reached $R_{\mathrm{exact}} = 2.948572$, surpassing the formal leaderboard target.
\end{itemize}

\subsection{Topology Comparison}

Table~\ref{tab:topologies} compares the seed topologies tested against the best results found.

\begin{table}[t]
\centering
\caption{Comparison of seed topologies.}
\label{tab:topologies}
\begin{tabular}{@{}lrl@{}}
\toprule
Topology & Best $R$ & Notes \\
\midrule
3-11-1 (current best) & 2.948572 & Found by fast MCMC + nano polish \\
4-11 ring (prior best) & 2.960344 & Found by LNS3 from SA seed \\
C5 pentagonal ($5\times3$) & 3.453 & 5 groups of 3 \\
6 conjugate pairs + 3 singles & 3.510 & Antiparallel flat-face pairs \\
Brickwall random & 3.516 & Grid, random orientations \\
7 conjugate pairs + 1 single & 3.582 & Low valid-seed rate (6/20) \\
C5 loose & 3.566 & Wider C5 grouping \\
Brickwall tight & 3.609 & $1.8 \times 1.6$ grid spacing \\
Brickwall standard & 3.723 & $2.1 \times 2.0$ grid spacing \\
\bottomrule
\end{tabular}
\end{table}

The 4-11 and 3-11-1 topologies both strongly outperform all others, but they are structurally distinct basins: direct perturbation from 4-11 configurations could not reach 3-11-1 geometries.

% ================================================================
\section{Analysis}
\label{sec:analysis}

\subsection{Why Sophisticated Methods Failed}

\paragraph{L-BFGS-B with phi-function energy.}
This consistently converged to solutions worse than the greedy hill-climber.
The root cause is the $\Phi_{PP}$ component of Equation~\eqref{eq:phiPP}: it is only valid when flat-face normals are anti-parallel ($\mathbf{n}_i \cdot \mathbf{n}_j \le -1 + \varepsilon$).
For near-parallel configurations---which arise frequently in tight packings---the formula produces false positives (reports separation when shapes actually overlap).
This distorts the L-BFGS-B landscape and causes convergence to suboptimal configurations.

\paragraph{Monotonic basin hopping (MBH).}
Following \citet{grosso2010}, we implemented strict monotonic acceptance with L-BFGS-B as the local minimiser.
After 1000+ rounds, zero improvement over baseline.
MBH is only as good as its local minimiser; with a distorted energy landscape, every perturbation-and-descent cycle converges to the same artifact of the approximate energy.

\paragraph{Population basin hopping (PBH).}
Same root cause as MBH. Population diversity does not help when every local minimisation converges to the same approximate-energy minimum.

\paragraph{Formulation space search (FSS).}
Converting to polar coordinates $(r, \varphi, \theta)$ changes the optimisation geometry but not the underlying energy-function approximation.

\paragraph{Orientation-flip search.}
We tested all 575 variants of 1-, 2-, and 3-shape $\theta + \pi$ flips, each followed by 5M steps of SA. Zero new bests. The leaderboard cluster at $R \approx 2.9727$ is not accessible from our configuration via orientation flips---it is in a structurally different basin.

\paragraph{Random multistart.}
50 independent SA runs from random starts (50M steps each) yielded $R \approx 3.06$ at best. Random configurations rarely produce the tightly packed ring topology.

\subsection{The Basin Discovery Problem}
\label{sec:landscape}

Our best solution topologies were found through two distinct mechanisms: serendipitous topological collapse (SA falling into a 4-11 arrangement) and high-throughput approximate exploration (fast MCMC exploring the 3-11-1 region). Neither was achievable by careful exact refinement within the prior basin.

This pattern suggests the landscape for semicircle packing is multi-modal with high inter-basin barriers. The barriers appear to require a minimum contact-graph change to cross, making them inaccessible to continuous exact-oracle perturbations that preserve local contacts.

\subsection{Oracle Cost Determines Method Choice}
\label{sec:oraclecost}

This study provides an extended case study where the cost of the feasibility oracle dominated method performance:

\begin{itemize}
  \item Numba SA (${\sim}1$M steps/sec): found $R = 2.976$.
  \item Greedy hill-climber (${\sim}120$ eval/sec, exact Shapely): found $R = 2.974679$.
  \item MBH/PBH + L-BFGS-B (${\sim}100$ iter/hr): found nothing.
  \item LNS3 with GJK polish (${\sim}12$ cycles/min): $\Delta R = 0.013$ in 91 seconds.
  \item Fast approximate MCMC (${\sim}300$k batches/run, Shapely gate): $\Delta R = 0.011$ in one run.
\end{itemize}

The pattern is instructive: approximate fast methods dominated for large-scale basin exploration, while exact methods dominated for local refinement. Neither alone was sufficient to reach the current best.

We propose the following heuristic for method selection:

\begin{itemize}
  \item Oracle cost $< 1$\,ms: greedy hill-climbing or SA with exact evaluation at every step.
  \item Oracle cost 1--100\,ms: fast approximate inner loop with exact validation gate at save events.
  \item Oracle cost $> 100$\,ms: surrogate models or population methods; approximate energy throughout.
\end{itemize}

\subsection{Approximate vs.\ Exact Score Gap}

A persistent feature of the fast MCMC phase is a systematic gap between approximate and exact scores.
Valid solutions with approximate score $R_{\mathrm{approx}} = 2.948586$ consistently validated to exact scores in the range $2.948911$--$2.948953$ before the nano polishing phase. The gap (${\sim}300$--$400$ millionths) arises from the cheap overlap proxy underestimating true distances. Subsequent nano polishing closed part of this gap, pushing the exact score from $2.948911$ to $2.948572$.

Twelve distinct configurations with approximate score $2.948586$ were known but failed exact validation due to marginal overlaps. Whether any of these can be nudged to validity without inflating the exact score remains an open question.

\subsection{Leaderboard Context}

As of our last leaderboard check (March 28, 2026): our score $R = 2.97468$ ranked \#9. A cluster at positions \#5--\#8 achieved $R \approx 2.9727$--$2.9728$. The leader achieved $R = 2.96175$.

Our current best ($R = 2.948572$, April 22, 2026) is approximately $0.013$ below the last-known leader, surpassing the formal target of $R < 2.94889$ by approximately 320 millionths.

% ================================================================
\section{Conclusions}
\label{sec:conclusions}

We have presented a computational study of packing 15 unit semicircles into a minimal enclosing circle, achieving a best result of $R = 2.948572$ after 27 days of computation (March 26 -- April 22, 2026) across multiple algorithmic phases and 662 git-tracked solution commits. Our principal findings are:

\begin{enumerate}
  \item \textbf{Ring topologies dominate.} The 3-11-1 topology (3 inner, 11 mid-ring, 1 boundary-defining) and 4-11 topology both strongly outperform all other tested configurations. They represent distinct basins separated by barriers that continuous exact-oracle methods could not cross.

  \item \textbf{Exact oracles beat approximate gradients for local refinement.} When feasibility can be evaluated exactly in ${\sim}0.5$\,ms, a greedy hill-climber using the exact oracle outperforms L-BFGS-B with approximate phi-function energy. The phi-function's $\Phi_{PP}$ component is unreliable for non-anti-parallel flat faces, distorting the optimisation landscape.

  \item \textbf{Approximate inner loops enable basin exploration.} The two largest score improvements ($\Delta R = 0.013$ via LNS, $\Delta R = 0.011$ via fast approximate MCMC) both required methods that could modify the contact graph or explore at speeds impossible for exact-oracle methods. Exact refinement alone, applied to either basin, could not reach the other.

  \item \textbf{The approximate/exact gap is a search signal.} Configurations with promising approximate scores that fail exact validation represent near-misses on the validity boundary. Retaining these and using them as restart seeds improved search diversity in the fast MCMC phase.

  \item \textbf{No conjugate pairs in the best solution.} Despite theoretical expectations, zero antiparallel flat-face contacts appear in the best solution, suggesting that for $N = 15$ the optimal regime avoids flat-face pair formation.
\end{enumerate}

\paragraph{Open questions.}
What is the true optimum for $N = 15$ (likely in $[2.85, 2.95]$ based on area bounds and observed landscape structure)? Does the global optimum use a ring topology, and if so which one? The 3-11-1 and 4-11 topologies are structurally distinct basins; whether a third, undiscovered topology exists below 2.948 is unknown. Can the phi-function be corrected for bounded semicircular segments to enable faster gradient-based polishing?

\paragraph{Reproducibility.}
Code is available at \url{https://github.com/CamArmstr/shape-packing-challenge}.
The best solution JSON (\texttt{best\_solution.json}), solutions archive (\texttt{solutions/}), and all optimiser implementations are included.

% ================================================================
\paragraph{Acknowledgements.}
Problem posed by benedictbrady via the shape-packing-challenge repository.
The fast approximate MCMC approach (Section~\ref{sec:fastmcmc}) was directly inspired by compusophy's browser-based parallel tempering optimizer \citep{compusophy2026}, whose explorer/polisher architecture and targeting of the 0.234 optimal acceptance rate informed our design.
Computational work performed on a Windows~11 / WSL2 machine with a 16-core Intel CPU.

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
